Methods for Analyzing Models of Rod-Shaped Bacteria

### 2025-07-18 paper/main.tex
formatting
extend descriptino of figure

\begin{figure}[H]
    \centering
    \tikzstyle{rect} = [
        rectangle,
        % rounded corners,
        minimum width=4cm,
        minimum height=1cm,
        text centered,
        draw=black,
        fill=color4!70,
    ]
    \tikzstyle{brect} = [rect, fill=color5!50]
    \tikzstyle{frect} = [rect, fill=color3!40]
    \tikzstyle{arrow} = [thick,->,>=stealth]
    \tikzstyle{darrow} = [arrow, dashed]
    \begin{tikzpicture}
        \node (model) [brect, align=center] at (0,0) {Mathematical \&\\ Computational Model};
        \node (extraction) [rect] at (0, -2) {Data Extraction};
        \node (params) [brect] at (6, -2) {Parameter Estimation};
        \node (visualization) [rect] at (6, 0) {Visualization};

@@

%###################################################################################################
\section{Method Description}
\subsection{Fundamental Assumptions on the Spatial Representation of Rod-Shaped Bacteria}
% TODO CITATIONS
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{v}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{v}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{v}_{i-1},\vec{v}_i$ can be used when modeling the
dynamics of these bacteria.

@@

%###################################################################################################
\subsection{Data Extraction}
\label{section:fitting-and-comparing-masks}

@@

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/image001042.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/mask-zoom.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.255\textwidth]
        {docs/source/_static/fitting-methods/algorithm/interpolate-positions.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}
    \caption{
        TODO change 2nd image!\\
        (A) An example of a microscopic image.
        (B) The result of the  skeletonization algorithm when applying it to a single cell.
        (C) Extrapolated vertices $\vec{v}_i$ which represent the position of our cell-agents.
    }
    \label{fig:position-extraction-algorithm}
\end{figure}

@@

%###################################################################################################
\subsection{Parameter Estimation}
In order to effectively estimate the parameters of our model, we need to set up a cost function
which can be minimized.
In doing so, we hope to identify all parameters, given relevant experimental data.
We use microscopic images and construct a novel algorithm that can extract the position of the agent
from a provided cell-mask.
These cell-masks are easily obtainable by either using existing segmentation tools or manually
annottating images.
Following this, we can define a cost-function and an optimization scheme to minimize said function.

@@

% --------------------------------------------------------------------------------------------------
\section{Application to Mechanics of Rod-Shaped Bacteria}
\label{subsubsection:mechanical-model-rod-shaped-bacteria}

>>>>

\begin{figure}[H]
    \centering
    \tikzstyle{rect} = [
        rectangle,
        % rounded corners,
        minimum width=4cm,
        minimum height=1cm,
        text centered,
        fill=color4!70,
    ]
    \tikzstyle{brect} = [rect, fill=color5!50]
    \tikzstyle{frect} = [rect, fill=color3!40]
    \tikzstyle{arrow} = [thick,->,>=stealth]
    \tikzstyle{darrow} = [arrow, dashed]
    \begin{tikzpicture}
        \node (model) [brect, align=center] at (0,0) {Mathematical \&\\ Computational Model};
        \node (extraction) [rect] at (0, -2) {Data Extraction};
        \node (params) [brect] at (6, -2) {Parameter Estimation};
        \node (visualization) [rect] at (6, 0) {Visualization};

@@

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Fundamental Assumptions on the Spatial Representation of Rod-Shaped Bacteria}
% TODO CITATIONS
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{v}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{v}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{v}_{i-1},\vec{v}_i$ can be used when modeling the
dynamics of these bacteria.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Data Extraction}
\label{section:fitting-and-comparing-masks}

@@

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/image001042.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/mask-zoom.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.255\textwidth]
        {docs/source/_static/fitting-methods/algorithm/interpolate-positions.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}
    \caption{
        (A) Input image.
        (B) Result of the skeletonization algorithm for a single cell.
        (C) Vertices $\vec{v}_i$ which have been interpolated from the skeletonized cell.
        They represent the position of our cell-agents.
    }
    \label{fig:position-extraction-algorithm}
\end{figure}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Parameter Estimation}
In order to effectively estimate the parameters of our model, we need to set up a cost function
which can be minimized.
In doing so, we hope to identify all parameters, given relevant experimental data.
We use microscopic images and construct a novel algorithm that can extract the position of the agent
from a provided cell-mask.
These cell-masks are easily obtainable by either using existing segmentation tools or manually
annottating images.
Following this, we can define a cost-function and an optimization scheme to minimize said function.

@@

%###################################################################################################
\section{A Mechanical Model for Rod-Shaped Bacteria}
\label{subsubsection:mechanical-model-rod-shaped-bacteria}
